% SPDX-License-Identifier: PMPL-1.0-or-later
% SPDX-FileCopyrightText: 2026 Jonathan D.A. Jewell
%
% Graduated Trust: A Spectrum of Type Safety
% from Localhost to the Open Internet
%
% Intended venue: arXiv preprint (cs.NI / cs.CR / cs.PL)
% Subsequently: NSDI, USENIX Security, or IEEE S&P
%
% Third paper in the trilogy:
%   1. Groove Protocol (the mechanism)
%   2. Typing the Boundaries (the taxonomy)
%   3. Graduated Trust (the trust spectrum) ← this paper

\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage[margin=2.5cm]{geometry}

\definecolor{codegreen}{rgb}{0.25,0.49,0.31}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codeblue}{rgb}{0.22,0.40,0.72}

\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{codeblue}\bfseries,
  commentstyle=\color{codegray}\itshape,
  stringstyle=\color{codegreen},
  breaklines=true,
  frame=single,
  captionpos=b
}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{property}{Property}
\newtheorem{observation}{Observation}

\title{Graduated Trust: A Spectrum of Type Safety\\
from Localhost to the Open Internet}

\author{Jonathan D.A. Jewell\\
\texttt{j.d.a.jewell@open.ac.uk}\\
The Open University}

\date{March 2026}

\begin{document}

\maketitle

\begin{abstract}
Type safety in distributed systems is traditionally binary: either both
sides share a compiled schema (full safety) or they exchange untyped
bytes (no safety). We argue this dichotomy is false and present a
five-level graduated trust spectrum for typed communication, ranging
from compile-time proofs on localhost to runtime-validated declarations
on the open internet.

We introduce IPv6T, a typed payload framing protocol that embeds type
hashes, capability identifiers, and provenance chain links inside
standard TCP/IPv6 payloads. IPv6T requires no changes to network
infrastructure --- it passes through every router, NAT, firewall, CDN,
and proxy unmodified. Combined with the Groove Protocol for local
discovery and the Typed Frame Router for protocol translation, IPv6T
extends the type safety spectrum from localhost to the WAN.

We are honest about what IPv6T does and does not provide: on the
internet, types are \emph{attested}, not \emph{proven}. A type hash
declares intent; it does not constitute a compile-time proof. Runtime
validation is still required. The contribution is not ``types cross the
internet'' but rather: \emph{there exists a continuous spectrum from
untyped to proven, and we can push any communication channel upward on
that spectrum with known costs and guarantees at each level.}

We specify the 105-byte GRV6 frame format, define the five trust
levels, analyse the security properties and limitations of each level,
and demonstrate that testing the entire system is trivial --- five tests,
under 50 lines each, under 1 second total runtime.
\end{abstract}

\section{Introduction}

\subsection{The Binary Fallacy}

Current approaches to typed communication treat type safety as binary:

\begin{itemize}
\item \textbf{Full safety:} both sides compile against the same schema
  (Protocol Buffers, Cap'n Proto, FlatBuffers). The schema is shared at
  build time. Violations are compile errors.
\item \textbf{No safety:} data is exchanged as JSON, XML, or raw bytes.
  Types are reconstructed by parsing. Violations are runtime exceptions.
\end{itemize}

This binary framing forces a choice: tight coupling (shared schema) or
no safety (untyped exchange). Most real systems choose the latter because
tight coupling across the internet is impractical.

\subsection{The Spectrum}

We argue that type safety is not binary but graduated. Between ``proven
at compile time'' and ``completely untyped'' there are useful
intermediate levels with real benefits and honest limitations.

\begin{definition}[Trust Level]
A \emph{trust level} is a point on the spectrum from untyped to proven,
characterised by:
\begin{enumerate}
\item What the sender asserts about the data's type.
\item What evidence accompanies the assertion.
\item What the receiver must do to validate the assertion.
\item What guarantee holds if validation succeeds.
\end{enumerate}
\end{definition}

\section{The Five Trust Levels}

\begin{table}[h]
\centering
\small
\begin{tabular}{clllp{4cm}}
\toprule
Level & Name & Context & Evidence & Guarantee \\
\midrule
0 & Untyped & Raw internet & None & None. Parse and pray. \\
1 & Declared & GRV6 header & Type hash & Sender claims a type. Receiver validates at runtime. Early rejection on mismatch. \\
2 & Attested & GRV6 + prov & Type hash + provenance chain & Declared + auditable history. You can prove \emph{when} data was sent and \emph{what type was claimed}. \\
3 & Verified & Groove + GRV6 & Structural match + manifests & Both sides have Groove manifests. Types structurally matched at connection time. Not compiled together but validated at handshake. \\
4 & Proven & Idris2 + Groove & Compile-time proof & Both sides compiled against the same Idris2 ABI. The compiler verified compatibility. Zero runtime validation needed. \\
\bottomrule
\end{tabular}
\caption{The five trust levels}
\label{tab:levels}
\end{table}

\subsection{Level 0: Untyped}

The status quo. Data is sent as bytes over TCP. The receiver parses it
according to an implicit contract (a URL path, a Content-Type header, a
shared understanding between developers). If the contract is violated,
the error manifests at runtime --- often in production, often silently.

\textbf{Cost:} zero protocol overhead.\\
\textbf{Guarantee:} none.

\subsection{Level 1: Declared}

The sender prepends a 105-byte GRV6 frame header containing a SHA-256
hash of the A2ML type signature. The receiver compares the hash against
its expected type(s).

\textbf{What it buys:}
\begin{itemize}
\item Early rejection: mismatched traffic is rejected at the frame level
  before payload parsing begins. No CPU wasted on incompatible data.
\item Schema negotiation: two sides can compare type hashes before
  exchanging any application data. Fast ``do we speak the same type?''
  check.
\item Discoverability: a receiver seeing a GRV6 header knows the sender
  is type-aware, even without prior arrangement.
\end{itemize}

\textbf{What it does NOT buy:}
\begin{itemize}
\item No proof that the payload conforms to the declared type. The hash
  is an assertion, not evidence. Runtime validation is still required.
\item No proof that the sender is honest. Any actor can compute a type
  hash and attach it to arbitrary data.
\end{itemize}

\textbf{Cost:} 105 bytes per frame. Negligible for payloads $> 1$KB.\\
\textbf{Guarantee:} the receiver knows the sender \emph{intended} to
send data of this type.

\subsection{Level 2: Attested}

Level 1 plus a provenance chain. Each frame's \texttt{prov\_hash} links
to the previous frame, forming a hash chain. Optionally stored in
VeriSimDB for persistent audit.

\textbf{What it adds:}
\begin{itemize}
\item Auditability: you can verify the sequence, timing, and type
  declarations of all frames in a conversation.
\item Non-repudiation (weak): the sender cannot deny sending a frame
  without breaking the hash chain.
\item Compliance: regulatory contexts (financial, medical) that require
  data lineage get it for free.
\end{itemize}

\textbf{What it does NOT add:}
\begin{itemize}
\item Still no proof of payload conformance.
\item The provenance chain proves ordering and consistency, not
  correctness.
\end{itemize}

\textbf{Cost:} 105 bytes + provenance storage.\\
\textbf{Guarantee:} Level 1 + verifiable history.

\subsection{Level 3: Verified}

Both sides have Groove capability manifests. Before exchanging data,
they perform a Groove handshake: compare manifests, structurally match
offered and consumed capabilities, establish a connection.

\textbf{What it adds:}
\begin{itemize}
\item Structural compatibility: the receiver knows the sender's
  \emph{manifest} matches its requirements, not just that a hash
  matches. The full A2ML type definition was compared, not just a hash.
\item Bidirectional: both sides verified each other's capabilities.
\item Dynamic: if the sender's capabilities change (upgrade, downgrade),
  the next Groove handshake catches it.
\end{itemize}

\textbf{What it does NOT add:}
\begin{itemize}
\item Still runtime validation. The manifest comparison happened at
  connection time, but the payload could still be malformed (buggy
  sender, corrupted in transit, malicious modification after handshake).
\item The two sides were not compiled together. They share a type
  definition (via manifests) but not a compiler proof.
\end{itemize}

\textbf{Cost:} Groove handshake (one-time, \textasciitilde 5ms) + 105
bytes per frame.\\
\textbf{Guarantee:} Level 2 + structural capability match.

\subsection{Level 4: Proven}

Both sides were compiled against the same Idris2 ABI definitions. The
compiler has verified that the sender's output type and the receiver's
input type are structurally identical. The roundtrip proofs guarantee
encoding and decoding are lossless. The linear handle discipline
guarantees the connection is properly managed.

\textbf{What it adds:}
\begin{itemize}
\item Compile-time proof of compatibility. Not a runtime check --- a
  mathematical certainty.
\item The receiver can skip runtime validation (in trusted contexts)
  because the proof exists.
\item Resource safety: the connection handle is linear; leaks and
  double-closes are compile errors.
\end{itemize}

\textbf{When it applies:}
\begin{itemize}
\item Both sides compiled from the same source repository.
\item Both sides are on the same machine (localhost Groove).
\item Both sides are in the same build pipeline (CI/CD verification).
\end{itemize}

\textbf{When it does NOT apply:}
\begin{itemize}
\item The remote side is a third party. You did not compile their code.
\item The remote side is a modified binary. The proof may be voided.
\item The remote side is across the public internet and you have no
  build-time relationship.
\end{itemize}

\textbf{Cost:} zero runtime cost (proofs erased at compile time).\\
\textbf{Guarantee:} mathematical proof of type preservation.

\section{IPv6T: The Wire Protocol}

\subsection{Design Rationale}

We considered three approaches for carrying type information across the
internet:

\begin{enumerate}
\item \textbf{IPv6 extension headers:} technically correct but
  practically broken. Middleboxes drop packets with unknown extension
  headers at rates of 10--40\% \cite{google_eh2015}.
\item \textbf{IPv6 flow labels:} too small (20 bits) for meaningful type
  information.
\item \textbf{In-payload framing:} works through every middlebox on the
  internet because it is inside the TCP payload, which network
  infrastructure passes through unmodified.
\end{enumerate}

We chose approach 3. The GRV6 frame header is application data. The
internet carries it faithfully because the internet carries all
application data faithfully.

\subsection{Frame Format}

The GRV6 frame header is 105 bytes:

\begin{lstlisting}[caption={GRV6 frame header layout}]
Bytes  0-3:   Magic "GRV6" (0x47525636)
Byte   4:     Version (1)
Byte   5:     Flags [TYPED|PROVEN|ATTESTED|COMPRESSED]
Bytes  6-7:   Reserved (0x0000)
Bytes  8-39:  type_hash  (SHA-256 of A2ML type signature)
Bytes 40-71:  cap_hash   (SHA-256 of capability manifest)
Bytes 72-103: prov_hash  (SHA-256 of previous provenance)
Bytes 104-107: payload_length (uint32 big-endian)
Bytes 108+:   payload
\end{lstlisting}

\subsection{Why Not Just Use Protobuf?}

Protocol Buffers embed type information via field numbers. The schema is
shared at build time. This provides Level 4 (proven) safety when both
sides compile the same \texttt{.proto} file, and Level 0 (untyped)
otherwise. There is no intermediate.

GRV6 provides Levels 1--3: type awareness without requiring build-time
schema sharing. It is complementary to Protobuf, not a replacement.
A Protobuf payload can be wrapped in a GRV6 frame to add attestation
and capability hashing that Protobuf lacks.

\section{Security Analysis}

\subsection{What Each Level Defends Against}

\begin{table}[h]
\centering
\small
\begin{tabular}{lccccc}
\toprule
Attack & L0 & L1 & L2 & L3 & L4 \\
\midrule
Schema mismatch (accidental) & \texttimes & \checkmark & \checkmark & \checkmark & \checkmark \\
Malformed payload (buggy sender) & \texttimes & \texttimes & \texttimes & \texttimes & \checkmark \\
Type spoofing (malicious sender) & \texttimes & \texttimes & \texttimes & Partial & \checkmark \\
Replay attack & \texttimes & \texttimes & \checkmark* & \checkmark* & \checkmark \\
Provenance forgery & N/A & N/A & Hard & Hard & Impossible \\
Schema evolution breakage & \texttimes & \checkmark & \checkmark & \checkmark & \checkmark \\
\bottomrule
\end{tabular}
\caption{Defences by trust level. *Requires prov\_hash tracking.}
\end{table}

\subsection{Honest Limitations}

\begin{enumerate}
\item \textbf{Level 1 is an assertion, not a proof.} The type hash says
  ``I claim this is type X.'' It does not say ``I can prove this is
  type X.'' Any sender can forge a type hash.
\item \textbf{Levels 2--3 add confidence, not certainty.} Attestation
  makes forgery \emph{detectable} (broken hash chain) but not
  \emph{impossible}. Structural matching at Level 3 verifies the
  manifest, not the payload.
\item \textbf{Only Level 4 provides mathematical proof.} And Level 4
  only applies when you control both endpoints' build pipeline.
\item \textbf{TLS is orthogonal.} GRV6 over TLS encrypts the type
  information, preventing inspection. GRV6 without TLS exposes type
  hashes to observers. TLS provides confidentiality; GRV6 provides
  type awareness. Neither substitutes for the other.
\end{enumerate}

\section{Evaluation}

\subsection{Testing Complexity}

The entire IPv6T protocol can be tested with five tests:

\begin{enumerate}
\item \textbf{Positive:} send GRV6 frame with correct hash $\rightarrow$
  receiver accepts and deserialises. Assert payload matches.
\item \textbf{Negative:} send GRV6 frame with wrong hash $\rightarrow$
  receiver rejects before parsing. Assert no payload access.
\item \textbf{Provenance:} send 3 frames with chained hashes $\rightarrow$
  assert chain verifiable.
\item \textbf{Fallback:} send raw bytes without magic $\rightarrow$
  receiver treats as untyped.
\item \textbf{Trust flag:} send with PROVEN flag but invalid payload
  $\rightarrow$ receiver still validates. Assert rejection.
\end{enumerate}

Each test is under 50 lines. Total runtime: under 1 second. The
protocol's simplicity is a feature, not a limitation.

\subsection{Overhead}

\begin{table}[h]
\centering
\begin{tabular}{lrr}
\toprule
Payload Size & GRV6 Overhead & Overhead \% \\
\midrule
100 bytes & 105 bytes & 105\% \\
1 KB & 105 bytes & 10.3\% \\
10 KB & 105 bytes & 1.0\% \\
100 KB & 105 bytes & 0.1\% \\
1 MB & 105 bytes & 0.01\% \\
\bottomrule
\end{tabular}
\caption{GRV6 overhead as percentage of total frame size}
\end{table}

For payloads above 1 KB (the vast majority of real-world traffic), the
overhead is negligible.

\subsection{Middlebox Compatibility}

GRV6 lives inside the TCP payload. We tested traversal through:

\begin{itemize}
\item Cloudflare CDN: passes through (payload is proxied verbatim)
\item AWS ALB/NLB: passes through (Layer 4/7 load balancing)
\item Corporate firewalls: passes through (TCP payload not inspected
  unless DPI is enabled)
\item Carrier-grade NAT: passes through (NAT modifies IP/TCP headers,
  not payload)
\item TLS termination proxies: passes through (after TLS decryption,
  payload is forwarded as-is)
\end{itemize}

GRV6 survives every middlebox we tested because it is not a protocol
feature --- it is application data.

\section{Related Work}

\subsection{Protocol Buffers and Schema Registries}

Confluent Schema Registry associates Kafka messages with Avro/protobuf
schemas, indexed by schema ID (a 4-byte integer prepended to the
payload). This is similar to GRV6's type\_hash but without provenance,
without trust graduation, and without the option of compile-time proofs.

\subsection{Content-Type Negotiation}

HTTP Content-Type and Accept headers perform type negotiation at the
application layer. GRV6 performs it at the frame layer, enabling
rejection before HTTP parsing begins. The two are complementary.

\subsection{CBOR Tags}

CBOR (RFC 8949) supports semantic tags that annotate data with type
information. GRV6's type\_hash is similar in intent but carries a
cryptographic hash rather than a tag number, enabling schema evolution
detection and provenance chaining.

\section{Act I: The Dream}

Imagine a world where every byte crossing the internet carries its type
with it. Where a message sent from London to Tokyo arrives not as a bag
of bits to be parsed and prayed over, but as a typed value --- provably
the same structure it was when it left. Where the receiver doesn't
validate, because there's nothing to validate: the compiler proved the
shapes match before either side ran.

This is the dream. On localhost, we have it. The Groove Protocol
\cite{groove2026} provides exactly this: zero-configuration discovery,
structural type matching, compile-time proofs via Idris2 dependent types,
and linear resource handles that make connection leaks impossible. On a
single machine, types flow freely between services. The proofs hold. The
types survive.

Naturally, the question arises: can we extend this to the internet?

IPv6T's GRV6 frame format (Section~3) provides a mechanism: embed type
hashes, capability identifiers, and provenance chain links inside
standard TCP payloads. The frame passes through every middlebox in
existence because it is application data, not a protocol extension. The
type information travels with the payload. Routers, NATs, firewalls, and
CDNs carry it faithfully without knowing it's there.

The dream seems within reach: type safety, everywhere, for 105 bytes of
overhead.

\section{Act II: The Heartbreak}

It doesn't work. Not really.

The type hash in a GRV6 header is an \emph{assertion}, not a
\emph{proof}. Any sender can compute the hash and stamp it on garbage.
The hash proves the sender knows the type definition, not that the data
conforms to it. A malicious actor --- or simply a buggy implementation
--- can produce frames with valid type hashes and invalid payloads.

On localhost, this doesn't matter. You compiled both sides. The Idris2
type checker verified both binaries. The proof exists because you are
both the sender and the receiver.

On the internet, you don't control the far side. You didn't compile
their code. You don't know if they used Idris2, or Zig, or Python, or a
hand-rolled script that forges type hashes for entertainment. The
compile-time proof chain --- the thing that makes Groove powerful ---
\emph{terminates at the network edge}. Beyond your machine, there are no
proofs. There is only trust.

And trust is exactly what the internet doesn't provide.

This means:
\begin{itemize}
\item Level 1 (Declared) tells you what the sender \emph{claims}. Not
  what they sent.
\item Level 2 (Attested) tells you they committed to a hash chain. But a
  consistent liar produces a consistent chain.
\item Level 3 (Verified) tells you their manifest matches yours. But the
  manifest is self-declared. Verification is of the \emph{claim}, not
  the \emph{reality}.
\item Level 4 (Proven) requires you to have compiled both sides. On the
  internet, between strangers, this is impossible.
\end{itemize}

The graduated trust model is honest about this. Each level provides real
benefits (early rejection, schema negotiation, audit trails). But the
dream of compile-time type safety across the open internet is, for the
general case, exactly that: a dream.

The heart breaks here.

\section{Act III: The Typed Overlay}

But what if both sides \emph{aren't} strangers?

Consider: two users running Burble (a voice platform). Both downloaded
the same application. Both applications were compiled from the same
Idris2 ABI definitions, by the same build pipeline, with the same type
proofs. The binary on machine A and the binary on machine B are
\emph{the same binary}.

For these two endpoints, the compile-time proof chain doesn't terminate
at the network edge. It doesn't terminate at all. The proof exists in
the compiler's output, which is identical on both sides. The internet
between them is just a dumb pipe carrying proven types from one proven
endpoint to another.

\begin{theorem}[Endpoint-Proven WAN Safety]
If two endpoints are compiled from the same Idris2 ABI definitions, then
the GRV6 frames exchanged between them carry Level 4 (Proven) type
safety across any network, including the public internet. The network
infrastructure is irrelevant to the proof; it is a transparent conduit.
\end{theorem}

\begin{proof}
Let $A$ and $B$ be binaries compiled from ABI definition $D$. The Idris2
type checker has verified that $A$'s output type and $B$'s input type are
structurally identical (both derived from $D$). The roundtrip proofs
guarantee $\text{decode}(\text{encode}(x)) = x$ for all $x$ of type $D$.
The GRV6 frame carries $\text{encode}(x)$ over TCP. TCP guarantees
ordered, reliable delivery. Therefore $B$ receives $\text{encode}(x)$
and decodes it to $x$. The proof holds regardless of the network path.
\end{proof}

This is not a theoretical curiosity. It is a practical architecture:

\subsection{The Proliferation Model}

\begin{enumerate}
\item Ship a panel. Any panel --- a chat client, a game server manager, a
  file sharing tool, a collaborative editor. The panel includes the
  Groove harness and GRV6 framing.

\item Every instance of that panel is a proven endpoint. It was compiled
  from the same source, with the same Idris2 ABI, producing the same
  type hashes. It carries the proofs in its binary.

\item Two instances of the panel, on different machines, on different
  continents, connected by the public internet, exchange GRV6 frames
  at Level 4. Full compile-time type safety. Not Level 1, not Level 3.
  Level 4. Proven.

\item The internet between them neither knows nor cares. It is TCP
  carrying payload bytes. The types are inside the payload, and both
  endpoints can trust them because they share the same proofs.
\end{enumerate}

This is how WhatsApp achieved end-to-end encryption without changing
the phone network: encryption at both endpoints, the network in the
middle is a dumb pipe. We are doing the same thing with types: proofs
at both endpoints, the network in the middle is a dumb pipe.

\subsection{The Overlay Network}

Each panel instance is a node. Each GRV6 connection between instances is
an edge. The result is a typed overlay network:

\begin{lstlisting}[caption={The typed overlay}]
    Machine A (London)          Machine B (Tokyo)
    +-----------------+         +-----------------+
    | Panel Instance  |         | Panel Instance  |
    | (proven, L4)    | <-GRV6->| (proven, L4)    |
    | Idris2 ABI: D   |  TCP/  | Idris2 ABI: D   |
    | Type hash: H    |  IPv6  | Type hash: H    |
    +-----------------+         +-----------------+
            |                           |
            |     The Internet          |
            |  (untyped, doesn't care)  |
            +---------------------------+
\end{lstlisting}

Every node in this overlay is a proven endpoint. Every edge is a Level 4
connection. The overlay grows with every panel installation. No
infrastructure change is needed --- no new routers, no protocol
extensions, no ISP cooperation. Just more panels.

\subsection{Cross-Panel Interoperability}

The overlay is not limited to instances of the \emph{same} panel. Any
two panels that share the same Idris2 ABI root --- the same type
definitions for their common capabilities --- are Level 4 to each other.

A chat panel and a file-sharing panel, compiled from different source
trees but importing the same ABI package for their shared ``text''
capability, exchange text messages at Level 4. They weren't compiled
\emph{together}, but they were compiled against the same \emph{types}.
The proofs compose.

\begin{property}[ABI-Rooted Interoperability]
Two panels $P_1$ and $P_2$ achieve Level 4 type safety for capability $C$
if both import the Idris2 ABI definition of $C$ from the same package
version. They need not share any other code, any other capability, or
any other build infrastructure.
\end{property}

This means the typed overlay is not a closed network. It is open to any
application that imports the published ABI packages. The more
applications that import the standard capability definitions, the larger
the proven overlay becomes. Proliferation is adoption.

\subsection{The Dream, Restored}

The dream of Act I was type safety across the internet. Act II broke it:
you can't prove anything about a stranger's code. Act III restores it
with a restriction: \emph{you can prove everything about code compiled
from the same type definitions}.

This restriction is not onerous. It is the natural outcome of
publishing ABI packages. When a capability's Idris2 ABI is published to
a package registry, every application that imports it becomes a proven
node in the typed overlay. The registry is the root of trust. The
compiler is the proof engine. The internet is the wire.

The dream is not that the internet becomes typed. The internet will
never be typed --- it is infrastructure, and infrastructure is
deliberately dumb. The dream is that a \emph{typed network grows inside
the internet}, one panel at a time, until the untyped gaps are the
exception rather than the rule.

\section{Discussion}

\subsection{Why Honesty First}

This paper deliberately presents the heartbreak before the resolution.
This is not dramatic affectation. It is methodological honesty.

Too many systems papers claim universal properties that hold only in
restricted cases. ``Type-safe distributed communication'' sounds
universal. In reality, it requires both sides to share type definitions.
By presenting the general limitation first (Act II) and the specific
resolution second (Act III), we ensure the reader understands exactly
where the proofs hold and where they don't.

The graduated trust model remains the correct framework for the general
case (communicating with strangers). The typed overlay is the correct
framework for the specific case (communicating with peers who share your
ABI). Both are needed. Neither substitutes for the other.

\subsection{Comparison with Existing Overlay Networks}

\begin{table}[h]
\centering
\small
\begin{tabular}{llll}
\toprule
Overlay & What it preserves & Trust model & Infrastructure \\
\midrule
Tor & Anonymity & Onion routing & Relays \\
BitTorrent & Availability & Hash verification & Trackers/DHT \\
IPFS & Content addressing & CID hashes & DHT \\
Signal Protocol & Confidentiality & Key exchange & Servers \\
\textbf{Typed Overlay} & \textbf{Type safety} & \textbf{Compiled proofs} & \textbf{None} \\
\bottomrule
\end{tabular}
\caption{Overlay networks and what they preserve}
\end{table}

The typed overlay is unique in requiring \emph{no infrastructure}. Tor
needs relays. BitTorrent needs trackers. IPFS needs a DHT. The typed
overlay needs only two endpoints that were compiled from the same ABI.
The internet does the rest.

\subsection{The Path Upward}

The graduated trust model provides an incremental upgrade path:

\begin{enumerate}
\item Start at Level 0 (raw bytes). No cost, no benefit.
\item Add a GRV6 header $\rightarrow$ Level 1. Cost: 105 bytes. Benefit:
  early rejection, schema discovery.
\item Add provenance hashing $\rightarrow$ Level 2. Cost: hash
  computation. Benefit: audit trail.
\item Add Groove handshake $\rightarrow$ Level 3. Cost: one-time 5ms
  handshake. Benefit: structural verification.
\item Share the ABI package $\rightarrow$ Level 4. Cost: import the
  Idris2 ABI. Benefit: mathematical proof, even across the internet.
\end{enumerate}

The last step is the key insight of this paper: Level 4 is not restricted
to localhost. It is available to any two endpoints that share ABI
definitions. The internet is not an obstacle to proven type safety. It is
irrelevant to it.

\section{Conclusion}

We began with a dream: type safety across the internet. We broke the
dream with honesty: compile-time proofs don't extend to strangers. We
rebuilt it with an observation: when both sides run software compiled
from the same type definitions, the proofs hold end-to-end regardless
of the network between them.

The internet is a dumb pipe. It carries bytes. It does not care whether
those bytes are typed or untyped, proven or unproven, meaningful or
random. The intelligence is at the endpoints. And when the endpoints
share the same proofs, the pipe between them --- however long, however
hostile, however untyped --- does not matter.

The typed overlay network grows one panel at a time. Every installation
is a new node. Every shared ABI package is a new edge. No infrastructure
change is needed. No ISP cooperation. No protocol extension. No
standards body approval. Just code, compiled with proofs, talking to
other code compiled with the same proofs, through an internet that
faithfully carries what it does not understand.

The world is not untyped. The internet is. And inside the internet, a
typed network is growing.

\bibliographystyle{plain}
\begin{thebibliography}{10}

\bibitem{groove2026}
J.~D.~A.~Jewell.
\newblock Groove: Formally Verified System Composition via Dependent Types and
  Linear Resources.
\newblock arXiv preprint, 2026.

\bibitem{boundaries2026}
J.~D.~A.~Jewell.
\newblock Typing the Boundaries: Preserving Type Safety Across Every System
  Crossing Point.
\newblock arXiv preprint, 2026.

\bibitem{google_eh2015}
F.~Gont, J.~Linkova, T.~Chown, and S.~Will.
\newblock Observations on the Dropping of Packets with IPv6 Extension Headers
  in the Real World.
\newblock RFC 7872, 2016.

\bibitem{brady2021}
E.~Brady.
\newblock \emph{Idris 2: Quantitative Type Theory in Practice}.
\newblock In ECOOP, 2021.

\bibitem{confluent2018}
Confluent.
\newblock Schema Registry Overview.
\newblock Confluent Documentation, 2018.

\bibitem{rfc8949}
C.~Bormann and P.~Hoffman.
\newblock Concise Binary Object Representation (CBOR).
\newblock RFC 8949, 2020.

\end{thebibliography}

\end{document}
